---
title: "Factors of Alcoholism Prevalance Rates"
author: "Krista Shuckerow"
date: "March 24, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We want to studey factors that affect alcoholism prevalence across the world. The World Health Organization has a variety of factors such as gender, region, break down of alcohol sales, the proportion of men and women who abstain from alcohol per country and region, which we can use in our analysis. The response variable is the prevalence of alcoholism in a country. We will collect data from the Global Health Observatory data repository which is maintained by the World Health Organization. 

A person is classified as suffering from alcoholism if they have a disorder attributable to the consumption of alcohol. This includes binge drinking, drinking to get drunk, and heavy excessive drinking, at least once in the past year. This strict definition of alcoholism is taken because of the health related issues alcohol can cause. Prevalence is defined as the percent of adults (15+ years of age) with a diagnosis of alcoholism during a given calendar year. All data comes from the year 2010.

The income index comes from the Human Development Index released by the United Nations in their human development reports.  The income index is defined as GNI per capita (2005 PPP International \$, using natural logarithm) expressed as an index using a minimum value of \$100 and a maximum value \$75,000.


# Analysis 

In this paper we will perform tests at a 95% confidence level unless otherwise stated. 

### Gender Differences

We will begin our analysis by determining if there is a significant difference in the proportion of men and women who suffer from alcoholism prevalence across the world. 

#### Raw Data

```{r echo = FALSE}
alcoholism <- read.csv("data.csv", header = F, row.names = 1)
colnames(alcoholism) <- as.character(unlist(alcoholism[2,]))
alcoholism <- alcoholism[c(-1,-2),]
alcoholism$Male <- gsub("\\[.{1,}", "", alcoholism$Male)
alcoholism$Female <- gsub("\\[.{1,}", "", alcoholism$Female)
alcoholism$`Both sexes` <- gsub("\\[.{1,}", "",alcoholism$`Both sexes`)
alcoholism$Male <- as.numeric(alcoholism$Male)
alcoholism$Female <- as.numeric(alcoholism$Female)
alcoholism$`Both sexes` <- as.numeric(alcoholism$`Both sexes`)

head(alcoholism)
```

Looking at the first few data points we see that the data is paired by country. Let's look at the distribution of the data. 


```{r echo=TRUE}
summary(alcoholism)
```
```{r echo = FALSE}
boxplot.default(alcoholism[c(-1, -2)], main = "Distribution of Proportion Suffering From Alcoholism", horizontal = T)

```

Each individual distribution appears skewed to the right. Both Male and Female distributions have various outliers. Looking at the summary of the data both the median and mean of the male distribution are higher than that of then women. 

#### Testing difference

We want to test the difference between the prevalence rate amoung men and women. We will use a paired t-test Our null hypothesis is there is no difference in the prevalence rates. Our alternative hypothesis is that male prevalence rates are higher than female prevalence rates. We will test at the 95% significance level. 

$$H_0: p_m - p_f \leq  0$$
$$H_a: p_m - p_f > 0 $$ 

```{r echo= TRUE}
difference <- alcoholism$Male - alcoholism$Female
summary(difference)

boxplot(difference, horizontal = TRUE, main = "Distribution in Difference of Alcoholism Prevalence")
```

Based on the summary statistics and the boxplot of the distribution of differences in alcoholism prevalence between males and females across the world does not appear normally distributed. Normality is an assumption of a paired t-test. We will preform a Shapiro-Wilk test of normality to be certain. 

```{r echo = FALSE}
shapiro.test(difference)
```

Looking at the result of the Shapiro-Wilk normality test we get a p-value of approximately zero, which is less than 0.05. Therefore, we will reject the null hypothesis of normality. We conclude the difference in alcohol prevalence is not normally distributed. Since the normality assumption is violatedwe instead use the non-parametric alternative to the paired t-test, the Wilcoxon signed-rank test, even thought sample size is large (n = 190).

We will test the null hypothesis that the mean rank difference in alcoholism prevalence between men and women is the same against the alternative hypothesis that the mean rank difference in alcoholism prevalence is greater in men. 

```{r echo=FALSE}
wilcox.test(difference, alternative = "greater", conf.int = T)
```

Based on the results of the Wilcoxon signed rank test we get a p-value of approximately zero. This is less that 0.05 therefore we concluded than males have a higher alcoholism prevalence rates. In addition we are 95% confident that the true difference is male and female alcoholism prevalance is 4.75 percentage points.

### Men & Women Correlation 

Even though men have higher alcoholism prevalence rate, there may be a relation between the prevalence rates of men and women in the same country. 

```{r echo = F}
plot(alcoholism$Male, alcoholism$Female, xlab = "male", ylab = "female", main = "Alcoholism Prevalence Rates")
```
Looking at the graph there appears to be a linear relation. So we'll try to fit a linear model. 

```{r echo = F}
cor.test(alcoholism$Male, alcoholism$Female)

```


Looking at the data we are 95% confident that there is between 80.50% and 88.52% correlation between the alcoholism prevalence rates of males and females.   

### Proportion of Alcohol Sales

We will now investigate how much a certain type of alcohol sold impacts alcoholism levels. We will create a regressive model where male and female are factors. We'll start with the effect of wine, to investigate if there is a relation between the proportion of alcohol sales in a country attributed to wine and spirits (hard liquor) and the prevalence of alcoholism. 

#### Raw Data 

```{r echo = F}
alcohol <- read.csv("purealcohol.csv", header = F, row.names = 1)
colnames(alcohol) <-  as.character(unlist(alcohol[2,]))
alcohol <- alcohol[c(-1, -2),]

# Formatting
alcohol$Spirits <- gsub("No data", "", alcohol$Spirits)
alcohol$Spirits <- as.numeric(alcohol$Spirits)
alcohol$Wine <- gsub("No data", "", alcohol$Wine)
alcohol$Wine <- as.numeric(alcohol$Wine)
alcohol$Beer <- gsub("No data", "", alcohol$Beer)
alcohol$Beer <- as.numeric(alcohol$Beer)

countries <- read.csv("data.csv",header = FALSE)
countries <- countries$V1
countries <- countries[c(-1,-2)]

countries2 <- c(countries, countries)
spirits<- c(alcohol$Spirits, alcohol$Spirits)
wine <- c(alcohol$Wine, alcohol$Wine)
beer <- c(alcohol$Beer, alcohol$Beer)
responses <- c(alcoholism$Male, alcoholism$Female)
factors <- c(rep("M", length(alcoholism$Male)), rep("F", length(alcoholism$Female)))

gender <- data.frame(countries2, responses, wine, spirits, factors)
colnames(gender) <- c("country", "response", "wine", "spirits", "factor")
gender <- gender[,-1]

head(gender)
```

The above is the first few rows of a data table containing the alcoholism prevalence levels of a specific country for a gender and the proportion of the countries alcohol sale that is wine. 


### Type of Alcohol
First, let's look at a scatterplot of the data. 

```{r echo = FALSE}
library(car)
scatterplot(gender$response ~ gender$wine + gender$factor,
            xlab = "Proportion of Alcohol Sold is Wine", 
            ylab = "Prevalence of Alcoholism", 
            main = "Alcoholism Prevalence and Wine Sales",
            legend = list(title = "Gender"))


scatterplot(gender$response ~ gender$spirits + gender$factor,
            xlab = "Proportion of Alcohol Sold is Spirits", 
            ylab = "Prevalence of Alcoholism", 
            main = "Alcoholism Prevalence and Spirits Sales",
            legend = list(title = "Gender"))
```


### Regional

Gender norms vary based on country, so we will include gender and region as variables that interact. 
```{r echo = F}
r <- c('Eastern Mediterranean', 'Europe', 'Africa', 'Americas' , 'Western Pacific', 'South-East Asia' )

#Start at the letter M
regions <- c(r[1], r[2], r[3], r[2], r[3], r[4], r[4], r[2], r[5], r[2], r[2], r[4], r[1], r[6], r[4], r[2], r[2], r[4], r[3], r[6], r[4], r[2], r[3], r[4], r[5], r[2], r[3], r[3], r[3], r[3], r[5], r[3], r[4], r[3], r[3], r[4], r[5], r[4], r[3], r[3], r[5], r[4], r[2], r[4], r[2], r[2], r[6], r[3], r[2], r[1], r[4], r[4], r[4], r[1], r[4], r[3], r[3], r[2], r[3], r[5], r[2], r[2], r[3], r[3], r[2], r[2], r[3], r[2], r[4], r[4], r[3], r[3], r[4], r[4], r[4], r[2], r[2], r[6], r[6], r[1], r[1], r[2], r[2], r[2], r[4], r[5], r[1], r[2], r[3], r[5], r[1], r[2], r[5], r[2], r[1], r[3], r[3], r[1], r[2], r[2], r[3], r[3], r[5], r[6], r[3], r[2], r[5], r[3], r[3], r[4], r[5], r[2], r[5], r[2], r[1], r[3], r[6], r[3], r[5], r[6], r[2], r[5], r[4], r[3], r[3], r[5], r[2], r[1], r[1], r[5], r[4], r[5], r[4], r[4], r[5], r[2], r[2], r[1], r[5], r[2], r[2], r[2], r[3], r[4], r[4], r[4], r[5], r[2], r[3], r[1], r[3], r[2], r[3], r[3], r[5], r[2], r[2], r[5], r[1], r[3], r[2], r[6], r[1], r[4], r[3], r[2], r[2], r[1], r[2], r[6], r[2], r[6], r[3], r[5], r[4], r[1], r[2], r[2], r[5], r[3], r[2], r[1], r[2], r[3], r[4], r[4], r[2], r[5], r[4], r[5], r[1], r[3], r[3])

classified <- cbind.data.frame(alcoholism, regions)
classified[,c(-1, -2)]

regions2 <- c(regions, regions)
data.2 <- cbind(gender, regions2)

par(mar=c(4,10,2,1))
region.boxplot <- boxplot(data.2$response~data.2$regions2, horizontal = T, las = 1, varwidth = T, xlab = "Alcoholism Prevalence Rates", col = c("blue", "red", "yellow", "green", "magenta", "orange"), main = "Prevalance Rates by Region", plot = TRUE)

```

### Income

Income in a country can also affect the ammount. We'll load in a data set from the United Nations' 2010 Human Development Report.
```{r echo = F}

GNI <- read.csv("income.csv", header = F)

income <- c(GNI$V2, GNI$V2)
data.3 <- cbind(data.2, income)
```


# Linear Regression Model 

Now we combine everything we've looked at up until this point. We'll create a linear regression model involving what kinds of alcohol a country buys, the standardized gross national income of the country, and gender with region. We'll assume interaction between gender and region because different genders have differnt rights across countries and sections of the world. 

```{r echo = F}
#data.3 has all values without NA omitted
data.4 <- na.omit(data.3)
lm.2 <- lm(response~wine + spirits + income + factor + regions2 , data = data.4)
summary(lm.2)
```
Model result:


Looking at summary of the model the only parameters that appear to have a significant impact are gender, income, and the interaction between eastern mediterranean and gender and interaction between gender and europe. Overal The model is significant with a p-value of approximately 0 and an R-squared value of 0.5907, meaning this model explains 59.07% of variaion in the prevalence rate of alcoholism.

We need to check the assumptions. 

```{r echo = FALSE}
plot(lm.2)
```
Looking at the residuals vs. fitted values we see a non constant variance problem, where variance appears to increase with larger values in a funnel pattern. The Normal Q-Q plot shows signs of a heavy tailed distribution and moderate skewedness to larger values.

We will try to fix the variance funneling by taking the log of the 


```{r echo = F}
log.response <- log(data.4$response)

log.response[!is.finite(log.response)] <- 0

lm.3 <- lm(log.response ~wine + spirits + income + factor * regions2, data = data.4)
summary(lm.3)

```
With the transformation to the log response we see some more significant variables. The p-value is close to zero indicating our model is statistically significant. We also have an increased R-squared value. The model explains 75.88% of the variation in alcohol prevalence around the world. Let's check the assumptions.

```{r echo = FALSE}
plot(lm.3)
```
The residual's versus fitted graph looks a lot better. The points are scattered more evenly around zero and appear to have normal variation. This transformation also has a better normal distribution. The left tail of the model does appear highly skewed. We could investigate by looking at the residuals. 


```{r echo = FALSE}
sort(lm.3$residuals)
```

Looking at the residual's we'll remove residual values with an absolute value greater than 2. 

```{r echo = F}
data.5 <- data.4[c(-362, -257, -265), ]
log.response2 <- log(data.5$response)

log.response2[!is.finite(log.response2)] <- 0

lm.5 <- lm(log.response2 ~wine + spirits + income + factor * regions2, data = data.5)
summary(lm.5)
plot(lm.5)

```

Removing the residuals didn't the problem. We'll try removing some of the variables from the model without the removed values and see if that helps make the problem more normal.


#Reduce Model
The proportion of alcohol consumed does not appear to be having much affect on the model. We will reduce the model and preform a partial F-test to see if the proportion of a country's expenditure on wine and alcohol has a significant effect on the model.

The null hypothesis is that wine and spirits have no effect on the regression model. The alternative is that there is some added effect on the model by at least one of those two variables.


```{r echo = F}
reduced.lm <- lm( log.response2 ~ income + factor + regions2 + factor * regions2, data = data.5)

anova(reduced.lm, lm.5)

```

Looking at the output we get a -value of 0.9806 which is larger than 0.05. We do not have enough evidence to reject the null hypothesis and we conclude that the proportion of wine and spirits sold in a country have no siginificant effect on the model. 


```{r echo = F}
summary(reduced.lm)
plot(reduced.lm)
```

Next, we want to see if certain regions have an effect on the model.  We'll look at an interaction plot of the data
```{r echo = F}
interaction.plot(data.5$factor, data.5$regions2, log.response2, col = c("red", "blue", "green", "black", "purple", "orange"))
```
For some reason there has to be text here
------------------------------------------


##Conclusion

We find that gender is the biggest factor in prevalence rates, and that men are more likely to suffer from alcoholism around the world than women. We also find that there is a positive relationship between the prevalence rate of men and women within a country, and the high prevalence rates of alcoholism in women then it's likely men will also experience a high prevalence rate of alcoholism. 

After we seperated gender out as a factor we analyzed the affect wine had on the model. Even thought our model identified wine as a useful regressor, the regression model it's self was inadequate because it violated the normality assumption deeming all the results useless. 



## TO DO
Health index -- This would be confounding becuase alcoholism is used to help define the health index
income ***  - Done
by region - DONE


level of education
